Typesafe Coordinate Systems in High-Throughput Sequencing Applications

### 2022-09-13 manuscript/main.tex
Minor tweaks for submission to arXiv.org
 - reordered OCaml, Python
 - fixed abstract grammatical error
 - Removed "Inter" font (arXiv doesn't have this despite in TeXLive 2020?)

\usepackage{fontspec}
%\setmainfont{CrimsonPro}
\setmainfont{Inter}
%\usepackage{ebgaramond-maths}
%\usepackage[bitstream-charter]{mathdesign}
%\usepackage{librebaskerville}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
High-throughput sequencing file formats and tools encode coordinate intervals with respect to a reference sequence in at least four distinct, incompatible ways. Integrating data from and moving data between different formats has the potential to introduce subtle off-by-one errors. Here, we introduce the notion of typesafe coordinates: coordinate intervals are not only an integer pair, but members of a type class comprising four types: the Cartesian product of a zero or one basis, and an open or closed interval end. By leveraging the type system of statically and strongly-typed, compiled languages we can provide static guarantees that an entire class of error is eliminated. We provide a reference implementation in D as part of a larger work (dhtslib), and a proofs of concept in Rust, Python, and OCaml. Exploratory implementations are available at \url{https://github.com/blachlylab/typesafe-coordinates}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Unfortunately, the use of alternative systems in different files and application programming interfaces (APIs) which must interoperate sets the stage for subtle off-by-one errors, which are a well-recognized class of problems, but also hard to avoid and sometimes difficult to detect.\cite{CWE193} We propose the use of compilers' type systems to aid bioinformatics software developers and to enforce correctness in a guaranteed way. By applying the principles we describe, the use of a statically typed, compiled programming language for the development of bioinformatics software has the potential to provide these guarantees essentially for free. We provide a reference implementation in D, and a proofs of concept in Rust, Python, and OCaml.

@@

Instead, consider strongly and statically typed languages with generic and polymorphic features. Define a type class \texttt{Interval} of which there are four types, corresponding to ZBHO, ZBC, OBHO, and OBC coordinate systems, and for which there exists a function, \textit{to}, (or conversely, \textit{from}) that maps between constituent types. Note that the types may be freely interconverted, as there exists a bijective mapping between any two coordinate systems (Figure \ref{fig:transitiongraph}). While these types could all be defined individually, one might instead leverage polymorphic features of the type system, especially in systems where a type can be parameterized on a constant value. 
\filbreak
In C++, D, or similar languages this could be done by templating a single \texttt{Interval} class or struct on some enums defining basis and openness:
\begin{verbatim}
    Interval <CoordSystem::zbho> i (0, 100);    // C++
    auto i = Interval!CoordSystem.zbho(0, 100); // D
    let i = Interval::<ZeroBased, HalfOpen>::from_int(0, 100); // rust
\end{verbatim}

@@

To instead create a function \texttt{fun} that accepts input of all coordinate systems, the function must be polymorphic or generic over the coordinate(s) type. Then in use, \texttt{fun} must be instantiated\footnote{In templated languages like C++, D, Rust the function would undergo monomorphization; in e.g. Haskell or OCaml not} with the desired coordinate system type and thereby be parameterized by a known \texttt{Coordinates} type, so type-safety is still achieved. In D this can be inferred, so the function will be automatically instantiated with the correct type. The coordinate system provides compile-time (via the type system) and run-time checks (bounds) to ensure coordinates are valid for their coordinate system; \textit{i.e.}, a one-based system must not contain 0 as a coordinate while a half-open coordinate system’s end coordinate must be greater than its start coordinate.

@@

\filbreak
A pseudocode\footnote{Currently we don't have a good gene-based query method in the GFF3 module, but we should! and, the current iteration of Interval does not encode contig name/id, but we could potentially add this via an \texttt{Optional!} field} excerpt is shown below.\\

\begin{lstlisting}[language=D,frame=single,basicstyle=\ttfamily\small]
// GFF3 files' coordinates are one-based closed
auto regionOfInterest = gff3file.genebounds("TP53");
// regionOfInterest is type Interval!CoordSystem.obc

@@

Both software engineering and bioinformatics are relatively newer disciplines among the applied sciences. In the same way that other engineering fields may apply and benefit from fundamental advances in mathematics and physics, the integrity, safety, and performance of software systems can be improved by applying lessons from computer science. These principles have also made their way to bioinformatics and computational biology, where authors have recognized the Rust language (\url{https://www.rust-lang.org}) for its memory safety attributes and its potential contributions to safer and more correct bioinformatics tools.\cite{koster_rust-bio_2016,perkel_why_2020}  Here, we demonstrate that leveraging the type system and compile-time guarantees of strongly typed languages like D, Rust, and others can reduce the error surface of software in the computational biology and bioinformatics space. These principles could also be extended to more loosely typed language such as Python, albeit with much weaker guarantees and a runtime overhead.

The reference implementation of Typesafe Coordinates as part of \textit{dhtslib}\cite{gregory_dhtslib_nodate} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D (\url{https://dlang.org/}), a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, Python, and OCaml, albeit with language-specific limitations in each of these.

@@

In conclusion, we have introduced the concept of Typesafe Coordinate Systems for bioinformatics and computational biology, and provided a reference implementation as part of a larger and more comprehensive high-throughput sequencing package \texttt{dhtslib}. We have also provided additional proofs-of-concept in various languages including Rust, an emerging language for bioinformatics. We hope that these safety concepts will be adopted by others and implemented either as standalone packages (leveraging macros and metaprogramming), or directly integrated into popular libraries and packages such as SeqAn,\cite{reinert_seqan_2017} BioPython,\cite{cock_biopython_2009}, Rust-Bio,\cite{koster_rust-bio_2016} and hts-nim.\cite{pedersen_hts-nim_2018} Using languages' type systems to eliminate error in specific problem domains is a powerful safety concept and should become routine in bioinformatics.

@@

The authors wish to thank Nicholas Wilson for critically reading the draft manuscript.

>>>>

%\usepackage{fontspec}
%\setmainfont{CrimsonPro}
%\setmainfont{Inter}
%\usepackage[sfdefault]{inter} %arXiv.org TeX live doesn't have inter.sty
%\usepackage{ebgaramond-maths}
%\usepackage[bitstream-charter]{mathdesign}
%\usepackage{librebaskerville}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
High-throughput sequencing file formats and tools encode coordinate intervals with respect to a reference sequence in at least four distinct, incompatible ways. Integrating data from and moving data between different formats has the potential to introduce subtle off-by-one errors. Here, we introduce the notion of typesafe coordinates: coordinate intervals are not only an integer pair, but members of a type class comprising four types: the Cartesian product of a zero or one basis, and an open or closed interval end. By leveraging the type system of statically and strongly-typed, compiled languages we can provide static guarantees that an entire class of error is eliminated. We provide a reference implementation in D as part of a larger work (dhtslib), and proofs of concept in Rust, OCaml, and Python. Exploratory implementations are available at \url{https://github.com/blachlylab/typesafe-coordinates}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Unfortunately, the use of alternative systems in different files and application programming interfaces (APIs) which must interoperate sets the stage for subtle off-by-one errors, which are a well-recognized class of problems, but also hard to avoid and sometimes difficult to detect.\cite{CWE193} We propose the use of compilers' type systems to aid bioinformatics software developers and to enforce correctness in a guaranteed way. By applying the principles we describe, the use of a statically typed, compiled programming language for the development of bioinformatics software has the potential to provide these guarantees essentially for free. We provide a reference implementation in D, and proofs of concept in Rust, OCaml, and Python.

@@

Instead, consider strongly and statically typed languages with generic and polymorphic features. Define a type class \texttt{Interval} of which there are four types, corresponding to ZBHO, ZBC, OBHO, and OBC coordinate systems, and for which there exists a function, \textit{to}, (or conversely, \textit{from}) that maps between constituent types. Note that the types may be freely interconverted, as there exists a bijective mapping between any two coordinate systems (Figure \ref{fig:transitiongraph}). While these types could all be defined individually, one might instead leverage polymorphic features of the type system, especially in systems where a type can be parameterized on a constant value. 
\filbreak
In C++, D, or similar languages this could be done by templating a single \texttt{Interval} class or struct on some enums defining basis and openness:
\begin{verbatim}
    // C++
    Interval <CoordSystem::zbho> i (0, 100);
    // D
    auto i = Interval!CoordSystem.zbho(0, 100);
    // Rust
    let i = Interval::<ZeroBased, HalfOpen>::from_int(0, 100);
\end{verbatim}

@@

To instead create a function \texttt{fun} that accepts input of all coordinate systems, the function must be polymorphic or generic over the coordinate(s) type. Then in use, \texttt{fun} must be instantiated\footnote{In C++, D, Rust the function would undergo monomorphization; in e.g. Haskell or OCaml not} with the desired coordinate system type and thereby be parameterized by a known \texttt{Coordinates} type, so type-safety is still achieved. In D this can be inferred, so the function will be automatically instantiated with the correct type. The coordinate system provides compile-time (via the type system) and run-time checks (bounds) to ensure coordinates are valid for their coordinate system; \textit{i.e.}, a one-based system must not contain 0 as a coordinate while a half-open coordinate system’s end coordinate must be greater than its start coordinate.

@@

\filbreak
A pseudocode excerpt is shown below.\\

\begin{lstlisting}[language=D,frame=single,basicstyle=\ttfamily\small]
// GFF3 files' coordinates are one-based closed
// regionOfInterest is type Interval!CoordSystem.obc
auto regionOfInterest = gff3file.genebounds("TP53");

@@

Software engineering and bioinformatics are relatively newer disciplines among the applied sciences. In the same way that other engineering fields may apply and benefit from fundamental advances in mathematics and physics, the integrity, safety, and performance of software systems can be improved by applying lessons from computer science. These principles have also made their way to bioinformatics and computational biology, where authors have recognized the Rust language (\url{https://www.rust-lang.org}) for its memory safety attributes and its potential contributions to safer and more correct bioinformatics tools.\cite{koster_rust-bio_2016,perkel_why_2020}  Here, we demonstrate that leveraging the type system and compile-time guarantees of strongly typed languages like D, Rust, and others can reduce the error surface of software in the computational biology and bioinformatics space. These principles could also be extended to more loosely typed language such as Python, albeit with much weaker guarantees and a runtime overhead.

The reference implementation of Typesafe Coordinates as part of \textit{dhtslib}\cite{gregory_dhtslib_nodate} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D (\url{https://dlang.org/}), a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, OCaml, and Python, albeit with language-specific limitations in each of these.

@@

In conclusion, we have introduced the concept of Typesafe Coordinate Systems for bioinformatics and computational biology, and provided a reference implementation as part of a larger and more comprehensive high-throughput sequencing package \texttt{dhtslib}. We have also provided additional proofs-of-concept in various languages including Rust, an emerging language for bioinformatics. We hope that these safety concepts will be adopted by others and implemented either as standalone packages (leveraging macros and metaprogramming), or directly integrated into popular libraries and packages such as SeqAn,\cite{reinert_seqan_2017} BioPython,\cite{cock_biopython_2009}, Rust-Bio,\cite{koster_rust-bio_2016} and hts-nim.\cite{pedersen_hts-nim_2018} Using type systems to eliminate error in specific problem domains is a powerful safety concept and should become routine in bioinformatics.

@@

The authors wish to thank Brent Pedersen and Nicholas Wilson for critically reading the draft manuscript, and Seb Mondet for reviewing the OCaml code.
